---
title: "Observing File Organisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Observing File Organisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Imagine importing a set of photographs from a camera, smartphone, or scanner.
Initially the files arrive as a flat collection with sequential filenames. During
review, a person organises them into meaningful folders representing subjects,
projects, or collections.

In this vignette we use nine tiny text files as stand-ins for photographs. The
example demonstrates how `fscontext` records observations before and after this
organisation, allowing the contextual changes to be reconstructed

```{r setup}
library(fscontext)
library(dplyr)

vignette_folder <- file.path(tempdir(), "vignette")
camera <- file.path(vignette_folder, "camera")
photos <- file.path(vignette_folder, "photos")

unlink(c(vignette_folder, camera, photos), recursive = TRUE)

dir.create(vignette_folder)
dir.create(camera)
dir.create(photos)

for (i in 1:9) {
  writeLines(as.character(i),file.path(camera, paste0("P", i, ".txt")))
}

```


```{r showcamera}
dir(camera)
```
Let's take a snapshot as the files arrive from a camera or stored on a memory card.

```{r snapshotbefore}
before <- snapshot_storage(path = camera, 
                           root = vignette_folder, 
                           person_id = "photographer")
```

Summarise the files:

```{r summarybefore}
summary_before <- readRDS(before) |> 
  dplyr::select(person_id, rel_path, stem, size)

print(summary_before)
```

Now let's organise the photos into three groups: photos that are made inside of the house, in the garden, and photos that are not worth keeping and designated for deleting.

```{r organisephotos}
dir.create(file.path(photos, "house"), recursive = TRUE)
dir.create(file.path(photos, "garden"))
dir.create(file.path(photos, "delete"))

file.rename(
  file.path(camera, c("P1.txt","P2.txt","P5.txt")),
  file.path(photos, "house",
            c("P1.txt","P2.txt","P5.txt"))
)

file.rename(
  file.path(camera,
            c("P3.txt","P4.txt","P6.txt","P7.txt")),
  file.path(photos, "garden",
            c("P3.txt","P4.txt","P6.txt","P7.txt"))
)

file.rename(
  file.path(camera,
            c("P8.txt","P9.txt")),
  file.path(photos, "delete",
            c("P8.txt","P9.txt"))
)
```
The resulting file structure:

```text
photos/

    house/
        P1.txt
        P2.txt
        P5.txt

    garden/
        P3.txt
        P4.txt
        P6.txt
        P7.txt

    delete/
        P8.txt
        P9.txt
```

The second snapshot records the reorganised collection. Although the files
themselves have not changed, their contextual location within the filesystem has.

```{r after}
after <- snapshot_storage(path = photos, 
                          root = vignette_folder, 
                          person_id = "archivist")
```

We can see the new summary:

```{r summaryafter}
summary_after <- readRDS(after) |> 
  dplyr::select(person_id, rel_path, stem, size)

print(summary_after)
```

Now compare the two observations.

The files are matched using their filename (stem) and file size. The
comparison reveals how each file moved from the original camera import into its
new contextual location.

```{r joinsummaries}

summary_path_change <- summary_before |>
  rename(
    rel_path_before = rel_path,
    person_before = person_id
  ) |>
  left_join(
    summary_after |>
      rename(
        rel_path_after = rel_path,
        person_after = person_id
      ),
    by = c("stem", "size")
  ) |>
  select(
    stem, size,
    person_before, rel_path_before,
    rel_path_after, person_after
  ) |>
  filter(grepl("txt$", rel_path_before))


print(summary_path_change)
```

For example, photograph P1 was originally observed in the camera import and
later observed in the house folder.
```{r}
summary_path_change |>
  filter(stem == "P1") |>
  select(
    person_before,
    rel_path_before,
    rel_path_after,
    person_after
  ) |>
  tidyr::unite(
    col = "transition",
    rel_path_before,
    rel_path_after,
    sep = " → "
  )
```
The transition

`camera/P1.txt → photos/house/P1.txt` 

records an observable change in filesystem context.

From the perspective of `fscontext`, this is evidence that the file became part
of the contextual group house. The package deliberately does _not_ infer why
this happened. The folder may represent the subject of the photograph, a
temporary workspace, or another organisational decision known only to the user.

The important point is that the filesystem itself records a human curation
activity. By comparing two observational snapshots, `fscontext` reconstructs
this activity as reproducible evidence.

The resulting comparison can serve as the starting point for later semantic
review. A subsequent workflow may interpret the movement into `house`,
`garden`, or `delete` as candidate semantic assertions, record the provenance of
those assertions, and subject them to human review before they become part of a
knowledge graph or archival description.