---
title: "recordset_df"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{recordset_df}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to construct a semantically annotated `recordset_df` from ordinary filesystem observations. As a minimal example, we create five files representing two Records. Record `a` consists of a textual description and an image of a museum object. Record `b` consists of a textual description, a digital surrogate, and an OCR transcription of an archival document.

```{r setup}
library(fscontext)
tmp_dir <- file.path(tempdir(), "recordset_df")
if (dir.exists(tmp_dir)) unlink(tmp_dir, recursive = TRUE)
dir.create(tmp_dir)

writeLines(
  c("<html>", "<body>", "Description of the A object.", "</body>", "</html>"),
  file.path(tmp_dir, "a.html")
)
Sys.sleep(1)

writeBin(charToRaw("JPEG"), file.path(tmp_dir, "a.jpg"))
Sys.sleep(1)

writeLines(
  c("<html>", "<body>", "Description of the B document.", "</body>", "</html>"),
  file.path(tmp_dir, "b.html")
)
Sys.sleep(1)

writeLines(
  c("%PDF-1.4", "Digital surrogate of B."),
  file.path(tmp_dir, "b.pdf")
)
Sys.sleep(1)

writeLines(
  "Plain text transcription of B.",
  file.path(tmp_dir, "b.txt")
)
```

We observe the temporary directory using `snapshot_storage()`. The resulting snapshot records file-level observations such as names, timestamps, checksums and other filesystem metadata without making any assumptions about the semantic relationships between the files.

```{r makesnapshot}
rs001_snapshot_file <- snapshot_storage(
  path = tmp_dir,
  root = tmp_dir
)
```

Next we add simple curatorial metadata. In this example we assign a human-readable description to each observed file and indicate which files belong to the same Record. You can add any further metadata or data about the records.

```{r subsetsnapshot}
rs001_snapshot <- readRDS(rs001_snapshot_file)
rs001_snapshot$description <- c(
  "Description of Object A", "Image of Object A",
  "Description of Record B", "Surrogate of Record B", "OCR Text of Record B"
)
rs001_df <- rs001_snapshot[
  ,
  c("stem", "filename", "description", "quick_sig", "ctime")
]
```

## recordset_df

The recordset_df class extends `dataset::dataset_df` with lightweight semantics for describing Record Sets, Records and Record Parts. Rather than implementing the complete RiC ontology, it provides a small number of conventions that support reproducible workflows while remaining compatible with ordinary tidy data.

The`recordset_df` uses `dataset_df` internally for metadata, provenance and serialisation, which is an extended `tibble::tibble()` `tbl_df` data frame. Users who require richer metadata or publication-oriented functionality can use the methods provided by the [dataset package](https://dataset.dataobservatory.eu/articles/dataset_df.html) directly.

```{r createrecordset}
rs001 <- recordset_df(
  x = rs001_df,
  creator = utils::person("Jane", "Doe", role = "aut"),
  title = "Demonstrator Record Set",
  record_set_identifier = "http://example.com/archive/sets/rs001",
  description = "A demonstration of a record set",
  record_identifier = "stem",
  record_part_identifier = "filename"
)
```

The constructor warns that the Record identifiers are not unique. This is expected because each Record is represented by multiple observed files. In this example, Record `a` has two Record Parts (individual files) and Record `b` has three Record Parts (files), so the Record identifier necessarily occurs more than once.

```{r printrecordset}
print(rs001)
```

The `stem` column is declared to contain identifiers of RiC Records. The values are annotated as `rico:Identifier` objects and labelled "Record Identifier", allowing downstream software to distinguish Record identifiers from other identifiers without requiring a complete RiC knowledge graph. (See: [rico:Record](https://www.ica.org/standards/RiC/RiC-O_1-0-2.html#Record))

```{r recordlevel}
rs001$stem
```

The `filename` column is declared to contain identifiers of RiC Record Parts. Record `a` consists of two Record Parts—a textual description and an image of the object—while Record `b` consists of three Record Parts: a textual description, a digital surrogate and an OCR transcription. Each file therefore identifies an individual Record Part within its parent Record. (See: [rico:RecordPart](https://www.ica.org/standards/RiC/RiC-O_1-0-2.html#RecordPart))

```{r recorpart}
rs001$filename
```

This example illustrates the intended role of `recordset_df`: observational evidence is acquired first, and lightweight semantic assertions are added afterwards. The resulting object remains an ordinary `data.frame` while carrying sufficient metadata to support reproducible archival, curatorial and semantic enrichment workflows.
